---
title: "scRNA-seq Analysis of PBMC 3k Dataset"
author: "Dhvani Solanki"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
---

title: "scRNA-seq Analysis of PBMC 3k Dataset"
author: "Dhwanit"
date: "`r Sys.Date()`"
output: html_document
---------------------

## Overview

This project performs an end-to-end single-cell RNA-seq analysis of the PBMC 3k dataset using Seurat.
The workflow includes quality control, normalization, dimensionality reduction, clustering, marker identification, and cell type annotation.

---

## 1. Load Libraries

```{r}
library(Seurat)
library(tidyverse)
library(cowplot)
library(ggplot2)
```

---

## 2. Load Data

```{r}
pbmc.data <- Read10X(data.dir = "data/filtered_gene_bc_matrices/hg19/")

pbmc <- CreateSeuratObject(counts = pbmc.data,
                           project = "pbmc3k",
                           min.cells = 3,
                           min.features = 200)

pbmc
```

---

## 3. Quality Control (QC)

```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

### Filter low-quality cells

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & 
                              nFeature_RNA < 2500 & 
                              percent.mt < 5)

pbmc
```

---

## 4. Normalization and Feature Selection

```{r}
pbmc <- NormalizeData(pbmc)

pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 1500)
```

---

## 5. Scaling and PCA

```{r}
pbmc <- ScaleData(pbmc)

pbmc <- RunPCA(pbmc)

ElbowPlot(pbmc)
```

---

## 6. Clustering

```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:8)

pbmc <- FindClusters(pbmc, resolution = 0.5)
```

---

## 7. UMAP Visualization

```{r}
pbmc <- RunUMAP(pbmc, dims = 1:8)

DimPlot(pbmc, reduction = "umap", label = TRUE)
```

---

## 8. Marker Gene Identification

```{r}
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE,
                               min.pct = 0.3,
                               logfc.threshold = 0.3)

head(pbmc.markers)
```

### Top markers per cluster

```{r}
top10 <- pbmc.markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)

top10
```

---

## 9. Heatmap of Marker Genes

```{r}
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

---

## 10. Cell Type Annotation

```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T",
                     "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")

names(new.cluster.ids) <- levels(pbmc)

pbmc <- RenameIdents(pbmc, new.cluster.ids)
```

### UMAP with cell type labels

```{r}
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5)
```

---

## 11. Differential Expression Analysis

```{r}
mono.de <- FindMarkers(pbmc, ident.1 = "CD14+ Mono", min.pct = 0.25)

head(mono.de, 10)
```

---

## 12. Key Findings

* ~2600 cells retained after quality control filtering
* 9 distinct immune cell populations identified
* Clear separation of clusters in UMAP space
* Marker genes align with known PBMC biology

---

## 13. Session Info

```{r}
sessionInfo()
```
